Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})

### 2014-05-17 manuscript.tex
edits to discussion

%TO DO
%Fix figures and tables
%check citations in last paragraph for ter1
%reformat for AJB
%preprint somewhere
%finish fixing paragraph that jeff doesn't like
%check map details
%make figs smaller so they don't go to end of document
%Code for SAS and command line stuff on Git
%Phenotyping and genotyping data on Git (it's on Figshare?)

@@

Adaptation occurs either due to selection on standing variation or on \emph{de novo} mutations. Adaptation as a result of selection on standing variation has been well-described in a number of systems, for example, selection for lactose tolerance in humans \citep{Plantinga2012, Tishkoff2007}; variation at the \emph{Eda} locus in three-spined stickleback \citep{Kitano2008, Colosimo2005}; and pupal diapause in the Apple Maggot fly \citep{Feder2003}. Although the role of standing variation with respect to adaptation has been described in many systems, its importance to domestication is not as well studied.

In maize, alleles at important domestication loci (\emph{RAMOSA1}, \citep{SigmonVollbrecht2010}; \emph{barren stalk1}, \citep{Gallavotti2004}; and \emph{grassy tillers1}, \citep{Whipple2011}) have been shown to have been selected from standing variation, suggesting that diversity already present in teosinte may have played an important role in the domestication of maize. The \emph{teosinte branched1} gene has long been a central focus of research concerning maize domestication, and, while previous studies have suggested that differences in plant architecture between domesticated maize and teosinte are a result of selection on standing variation, little is known about variation at this locus in teosinte \citep{Clark2006, Studer2011}. \citet{Studer2011} genotyped 90 accessions of teosinte (inbred and outbred), providing the first evidence that the \emph{Hopscotch} insertion is segregating in teosinte \citep{Studer2011}.

Given that the \emph{Hopscotch} insertion has been estimated to predate the domestication of maize, it is not surprising that it can be found segregating in populations of teosinte. However, in sampling numerous individuals from many teosinte populations our study provides greater insight into the distribution and prevalence of the \emph{Hopscotch} in teosinte. While our findings are consistent with a previous study by \citet{Studer2011} in that we identified the \emph{Hopscotch} allele segregating in teosinte, we find it at higher frequency than previously suggested \citep{Studer2011}. Many of our populations with high frequency of the \emph{Hopscotch} allele fall in the Jalisco cluster identified by \citet{Fukunaga2005}, possibly suggesting a different history of the \emph{tb1} locus than in the Balsas region where maize was domesticated \citep{Matsuoka2002}. While gene flow from crops into their wild relatives is well-known, (\citep{Ellstrand1999, Zhang2009, Thurber2010, Baack2008, Hubner2012, Wilkes1977, VanHeerwaarden2011, Barrett1983}), our results are more consistent with \citet{Hufford2013} who found resistance to introgression from maize into teosinte \citep{Hufford2013}. Furthermore, \citet{Hufford2013} showed that domestication loci, such as \emph{tb1}, are particularly resistant to introgression in both directions of gene flow (i.e., maize to teosinte and teosinte to maize) \citep{Hufford2013}.

We find no evidence of recent introgression in our analyses. Clustering patterns in our NJ trees do not reflect a pattern expected if maize alleles at the \emph{tb1} locus had introgressed into populations of teosinte.  Moreover, analysis of linkage in the \emph{tb1} region does not reveal patterns of high LD relative to the rest of chromosome 1, and assignment to maize in this region in our STRUCTURE analysis is lower than the average across chromosome 1 (\ref{Fig3Structure}, Table \ref{Table4Q}). Together, these data point to an explanation other than recent introgression for the high observed frequency of \emph{Hopscotch} in some of  our \emph{parviglumis} populations.

Although recent introgression seems unlikely, we cannot rule out ancient introgression as an explanation for the presence of the \emph{Hopscotch} in these populations. If the \emph{Hopscotch} allele was introgressed in the distant past, they could have been sufficient recombination to break up any initial LD, leading to observations similar to those obtained here.  We find this scenario less plausible, however, as there is no reason why gene flow should have been high in the past but absent in present-day sympatric populations.  In fact, early generation maize-teosinte hybrids are easy to find in these populations today (MB Hufford, pers. observation), and genetic data support ongoing gene flow between domesticated maize and both \emph{Zea mays} ssp. \emph{mexicana} and \emph{Zea mays} ssp. \emph{parviglumis} in a number of sympatric populations \citep{Hufford2013, Ellstrand2007, vanHeerwaarden2011}.

Other explanation for differential frequencies of the \emph{Hopscotch} among teosinte populations include both drift and natural selection.  Previous studies using both SSRs and genome-wide SNP data have found evidence for a population bottleneck in the San Lorenzo population \citep{Hufford2010, Pyhajarvi2013}, and the lower levels of sequence diversity in the 5' UTR region and the more positive values of Tajima's D we present here are consistent with these findings. \jri{deviations from HWE may be consistent too if we see excess of homozygotes. do we?}. This bottleneck, however, does not explain differences in \emph{Hopscotch} allele frequency among populations, and the available information on diversity and population structure among these populations \citep{Hufford2010, Pyhajarvi2013} is not suggestive of colonization or other demographic events that might predict a high frequency of the allele in multiple populations.  \jri{here we need a few sentences on selection.  the 5' UTR has much more negative D than the upstream. do we know the Hop genotype for sequenced lines? can we separate the sequences into hop/no hop and look for differences? it wasn't until we did this that gt1 stuff really popped out.} \lev{we should know for some of them, i will check}
%   from expected genotype frequencies under Hardy-Weinberg equilibrium at the \emph{Hopscotch} insertion in the San Lorenzo population and additional populations may be indicative of demographic changes that have affected genome-wide patterns of variation, or may be the result of selective pressures acting on or near the \emph{tb1} locus..

@@

The phenotypic effects of the \emph{Hopscotch} insertion in domesticated maize have been well documented \citep{Clark2006, Studer2011}, and \citet{Weber2007} have described its effects in partially inbred lines of teosinte \citep{Weber2007} \jri{i don't think these were inbred.please doublecheck.}. Our study is the first to explicitly examine the phenotypic effects of the \emph{Hopscotch} insertion in individuals sampled from a natural population of teosinte. \jri{isn't this what weber did?? for 70+ populations!?} However, we found no significant effect of the \emph{Hopscotch} on tillering index or tiller number in our phenotyping experiments, and the effect of the \emph{Hopscotch} insertion in teosinte is discordant with that of maize. The lack of correlation between \emph{Hopscotch} genotype and tillering index or tiller number is surprising given its effects in maize. It is certainly possible that even though previous data demonstrate an effect of the \emph{Hopscotch} on tillering in maize \citep{Studer2011}, that the effect of the \emph{Hopscotch} in teosinte is more complicated and may be more difficult to observe. Moreover, \emph{tb1} is a single gene in a complex pathway that affects branching and tillering traits, and perhaps in combination with alleles at other loci the phenotypic effects of the \emph{Hopscotch} on tillering may not be consistent. \jri{thsi section still needs work. i think we have to do more here. weber shows an association between SNPs in tb1 and branch length. we need to discuss that!}

>>>>

%TO DO
%check citations in last paragraph for ter1
%reformat for AJB
%preprint somewhere
%finish fixing paragraph that jeff doesn't like
%check map details
%Code for SAS and command line stuff on Git
%Phenotyping and genotyping data on Git (it's on Figshare?)

@@

Adaptation occurs either due to selection on standing variation or on \emph{de novo} mutations. Adaptation as a result of selection on standing variation has been well-described in a number of systems; for example, selection for lactose tolerance in humans \citep{Plantinga2012, Tishkoff2007}, variation at the \emph{Eda} locus in three-spined stickleback \citep{Kitano2008, Colosimo2005}, and pupal diapause in the Apple Maggot fly \citep{Feder2003}. Although the role of standing variation with respect to adaptation has been described in many systems, its importance to domestication is not as well studied.

In maize, alleles at important domestication loci (\emph{RAMOSA1}, \citealt{SigmonVollbrecht2010}; \emph{barren stalk1}, \citealt{Gallavotti2004}; and \emph{grassy tillers1}, \citealt{Whipple2011}) have been shown to have been selected from standing variation, suggesting that diversity already present in teosinte may have played an important role in the domestication of maize. The \emph{teosinte branched1} gene has long been a central focus of research concerning maize domestication, and, while previous studies have suggested that differences in plant architecture between domesticated maize and teosinte are a result of selection on standing variation, little is known about variation at this locus in teosinte \citep{Clark2006, Studer2011}. \citet{Studer2011} genotyped 90 accessions of teosinte (inbred and outbred), providing the first evidence that the \emph{Hopscotch} insertion is segregating in teosinte \citep{Studer2011}.

Given that the \emph{Hopscotch} insertion has been estimated to predate the domestication of maize, it is not surprising that it can be found segregating in populations of teosinte. However, in sampling numerous individuals from many teosinte populations our study provides greater insight into the distribution and prevalence of the \emph{Hopscotch} in teosinte. While our findings are consistent with \citet{Studer2011} in that we identified the \emph{Hopscotch} allele segregating in teosinte, we find it at higher frequency than previously suggested \citep{Studer2011}. Many of our populations with high frequency of the \emph{Hopscotch} allele fall in the Jalisco cluster identified by \citet{Fukunaga2005}, possibly suggesting a different history of the \emph{tb1} locus than in the Balsas region where maize was domesticated \citep{Matsuoka2002}. While gene flow from crops into their wild relatives is well-known, \citep{Ellstrand1999, Zhang2009, Thurber2010, Baack2008, Hubner2012, Wilkes1977, VanHeerwaarden2011, Barrett1983}, our results are more consistent with \citet{Hufford2013} who found resistance to introgression from maize into teosinte \citep{Hufford2013}. Furthermore, \citet{Hufford2013} showed that domestication loci, such as \emph{tb1}, are particularly resistant to introgression in both directions of gene flow (i.e., maize to teosinte and teosinte to maize) \citep{Hufford2013}.

We find no evidence of recent introgression in our analyses. Clustering patterns in our NJ trees do not reflect a pattern expected if maize alleles at the \emph{tb1} locus had introgressed into populations of teosinte.  Moreover, analysis of linkage in the \emph{tb1} region does not reveal patterns of high LD relative to the rest of chromosome 1, and assignment to maize in this region in our STRUCTURE analysis is lower than the average across chromosome 1 (Fig. \ref{Fig3Structure}, Table \ref{Table4Q}). Together, these data point to an explanation other than recent introgression for the high observed frequency of \emph{Hopscotch} in some of  our \emph{parviglumis} populations.

Although recent introgression seems unlikely, we cannot rule out ancient introgression as an explanation for the presence of the \emph{Hopscotch} in these populations. If the \emph{Hopscotch} allele was introgressed in the distant past, recombination may have broken up LD, a pattern that would be consistent with our data.  We find this scenario less plausible, however, as there is no reason why gene flow should have been high in the past but absent in present-day sympatric populations.  In fact, early generation maize-teosinte hybrids are easy to find in these populations today (MB Hufford, pers. observation), and genetic data support ongoing gene flow between domesticated maize and both \emph{Zea mays} ssp. \emph{mexicana} and \emph{Zea mays} ssp. \emph{parviglumis} in a number of sympatric populations \citep{Hufford2013, Ellstrand2007, VanHeerwaarden2011}.

Additional explanations for differential frequencies of the \emph{Hopscotch} among teosinte populations include both drift and natural selection.  Previous studies using both SSRs and genome-wide SNP data have found evidence for a population bottleneck in the San Lorenzo population \citep{Hufford2010, Pyhajarvi2013}, and the lower levels of sequence diversity in the 5' UTR region and the more positive values of Tajima's D we present here are consistent with these findings. \jri{deviations from HWE may be consistent too if we see excess of homozygotes. do we?}. This bottleneck, however, does not explain differences in \emph{Hopscotch} allele frequency among multiple populations in the Jalisco group, and the available information on diversity and population structure among these populations \citep{Hufford2010, Pyhajarvi2013} is not suggestive of recent colonization or other demographic events that might predict a high frequency of the allele across populations.  \jri{here we need a few sentences on selection.  the 5' UTR has much more negative D than the upstream. do we know the Hop genotype for sequenced lines? can we separate the sequences into hop/no hop and look for differences? it wasn't until we did this that gt1 stuff really popped out.} \lev{we should know for some of them, i will check}
%   from expected genotype frequencies under Hardy-Weinberg equilibrium at the \emph{Hopscotch} insertion in the San Lorenzo population and additional populations may be indicative of demographic changes that have affected genome-wide patterns of variation, or may be the result of selective pressures acting on or near the \emph{tb1} locus..

@@

The phenotypic effects of the \emph{Hopscotch} insertion in domesticated maize have been well documented \citep{Clark2006, Studer2011}, and \citet{Weber2007} have described its effects in partially inbred lines of teosinte \citep{Weber2007} \jri{i don't think these were inbred.please doublecheck.}. Our study is the first to explicitly examine the phenotypic effects of the \emph{Hopscotch} insertion in individuals sampled from natural populations of teosinte. \jri{isn't this what weber did?? for 70+ populations!?} However, we found no significant effect of the \emph{Hopscotch} on tillering index or tiller number in our phenotyping experiments, and the effect of the \emph{Hopscotch} insertion in teosinte is discordant with that of maize. The lack of correlation between \emph{Hopscotch} genotype and tillering index or tiller number is surprising given its effects in maize. It is certainly possible that even though previous data demonstrate an effect of the \emph{Hopscotch} on tillering in maize \citep{Studer2011}, that the effect of the \emph{Hopscotch} in teosinte is more complicated and may be more difficult to observe. Moreover, \emph{tb1} is a single gene in a complex pathway that affects branching and tillering traits, and perhaps in combination with alleles at other loci the phenotypic effects of the \emph{Hopscotch} on tillering may not be consistent. \jri{thsi section still needs work. i think we have to do more here. weber shows an association between SNPs in tb1 and branch length. we need to discuss that!}
